Drivers of sex differences in consumption (NAMCS)

### 2017-06-28 sex-difference/namcs.Rmd
cleaned up sex diff

```{r load_data}
sex_codes = read_tsv('namcs_sex_codes.tsv') %>%
  rename(dx=code)

dx = read_tsv('../../db/namcs/data/namcs_2014_dx.tsv') %>%
  left_join(sex_codes, by='dx')

dx_visit = dx %>%
  group_by(visit_id) %>%
  summarize(uti='uti' %in% diagnosis_type,
            skin='skin' %in% diagnosis_type,
            vaginitis='vaginitis' %in% diagnosis_type)

top_abx = read_tsv('../../db/namcs/analysis/top_abx.tsv')

# abx_drug_codes = read_tsv('../../db/namcs/drug-codes/abx-codes.tsv') %>%
#   select(drug_code, antibiotic) %>%
#   mutate(antibiotic=tolower(antibiotic))

drug = read_tsv('../../db/namcs/data/namcs_2014_drugs.tsv')

abx = top_abx %>% left_join(drug, by='desc')

abx_visit = abx %>%
  select(visit_id, abx) %>%
  distinct() %>%
  mutate(value=TRUE) %>% spread(abx, value) %>%
  mutate(any_abx=TRUE)

visit = read_tsv('../../db/namcs/data/namcs_2014_visits.tsv') %>%
  left_join(dx_visit, by='visit_id') %>%
  replace_na(list(uti=FALSE, skin=FALSE, vaginitis=FALSE)) %>%
  left_join(abx_visit, by='visit_id') %>%
  mutate_at(c('any_abx', top_abx$abx), function(x) !is.na(x))
```

@@

## The most-consumed antibiotics

I wanted to focus my analysis on the top few most important antibiotics. The NAMCS data includes information about the therapeutic category of each drug. I looked through those categories and identified the ones that could refer to antibiotics.

```{r abx_cat_table}
abx_cat = read_tsv('../../db/namcs/data/parse/drug_codes/abx_codes.tsv')
abx_cat %>% kable
```

@@

```{r find_top_abx}
drug %>%
  semi_join(abx_cat, by=c('cat1'='category')) %>%
  select(visit_id, desc) %>%
  left_join(select(visit, visit_id, survey_id, stratum, weight), by='visit_id') %>%
  mutate(dummy=1) %>%
  svydesign(data=., ids=~survey_id, strata=~stratum, weights=~weight) %>%
  svyby(~dummy, ~desc, design=., svytotal) %>%
  arrange(desc(dummy)) %>%
  select(abx=desc, n_rx=dummy, se) %>%
  mutate(f=100*round(n_rx/sum(n_rx), 2),
         cum_f_rx=100*round(cumsum(n_rx/sum(n_rx)), 2),
         relative_se=100*round(se/n_rx, 2)) %>%
  select(drug=abx, `fraction (%)`=f, `relative SE (%)`=relative_se, `cumulative fraction (%)`=cum_f_rx) %>%
  head(15) %>%
  kable
```

I kept the drugs from azithromycin through clindamycin. I skipped moxifloxacin because it has a high relative SE and mostly won't make it through future analyses. I keep nitrofurantoin because it's interesting from the Medicare data.

## Overall consumption by sex

Overall, more of the antibiotic claims were made by women. Women consume roughly 21% more antibiotics than men.

```{r sex_total_odds}
drug %>%
  semi_join(abx_cat, by=c('cat1'='category')) %>%
  select(visit_id, desc) %>%
  left_join(select(visit, visit_id, sex, survey_id, stratum, weight), by='visit_id') %>%
  mutate(dummy=1) %>%
  svydesign(data=., ids=~survey_id, strata=~stratum, weights=~weight) %>%
  svyby(~dummy, ~sex, design=., svytotal) %>%
  mutate_at(vars(-sex), function(x) x/1e6) %>%
  mutate(fraction=dummy/sum(dummy),
         odds=fraction/min(fraction)) %>%
  select(sex, `no. rx. (millions)`=dummy, `SE (millions)`=se, odds) %>%
  kable
```

```{r models}
des = visit %>%
  mutate(sex=factor(sex, levels=c('male', 'female'))) %>%
  as.data.frame() %>%
  svydesign(data=., id=~survey_id, strata=~stratum, weights=~weight)

abx_total_f = function(abx) {
  svyby(as.formula(paste0('~', abx)), ~sex, des, svytotal) %>%
    rename_(.dots=setNames(list(paste0(abx, 'TRUE'), paste0('se.', abx, 'TRUE')), c('total', 'se'))) %>%
    select(sex, total, se) %>%
    mutate(abx=abx)
}

chisq_f = function(abx) {
  svychisq(as.formula(paste0('~sex+', abx)), des) %>%
    tidy %>%
    mutate(abx=abx) %>%
    select(abx, p.value)
}

ttest_f = function(abx) {
  svyttest(as.formula(paste0(abx, '~sex')), des) %>%
    tidy %>%
    mutate(abx=abx) %>% select(abx, estimate, p.value)
}

abx_totals = lapply(top_abx$abx, abx_total_f) %>% bind_rows()
abx_chisq = lapply(top_abx$abx, chisq_f) %>% bind_rows()
abx_ttest = lapply(top_abx$abx, ttest_f) %>% bind_rows()
```

For each drug, I asked how many visits were made, stratified by sex. I show two lines on the graph. The lower lines is exactly $y=x$ and shows the line of equal consumption. The upper line has a slope equal to the odds of a visit belong to a female; this is the line of equal consumption per visit.

```{r total_plot}
abx_totals %>%
  mutate_at(vars(total, se), function(x) x/1e6) %>%
  mutate(ci=se*1.96,
         ci_low=total-ci,
         ci_high=total+ci) %>%
  select(abx, sex, total, ci_low, ci_high) %>%
  gather('key', 'value', total:ci_high) %>%
  mutate(key=case_when(.$sex=='male' & .$key=='total' ~ 'x',
                       .$sex=='male' & .$key=='ci_low' ~ 'xmin',
                       .$sex=='male' & .$key=='ci_high' ~ 'xmax',
                       .$sex=='female' & .$key=='total' ~ 'y',
                       .$sex=='female' & .$key=='ci_low' ~ 'ymin',
                       .$sex=='female' & .$key=='ci_high' ~ 'ymax')) %>%
  select(-sex) %>%
  spread(key, value) %>%
  mutate(ymin=if_else(ymin<0, 0, ymin)) %>% # fix the negative ymin for moxi
  mutate(highlight=abx %in% c('nitro', 'amox', 'amoxclav')) %>%
  ggplot(aes(x=x, xmin=xmin, xmax=xmax, y=y, ymin=ymin, ymax=ymax, label=abx, color=highlight)) +
  geom_abline(col='gray') + geom_abline(slope=1.21277, col='gray') +
  geom_point() + geom_errorbar() + geom_errorbarh() +
  scale_color_manual(values=c('black', 'red')) +
  geom_text(nudge_x=1, nudge_y=0.5) +
  scale_x_continuous(breaks=c(0, 5, 10), limits=c(0, 10)) + xlab('No. rx. for men (millions)') +
  scale_y_continuous(breaks=c(0, 5, 10), limits=c(0, 12)) + ylab('No. rx. for women (millions)') +
  coord_fixed() +
  theme_minimal()
```

Caveat: As per NAMCS recommendations, the male/nitro combination is too small (20 records) and too disperse (relative standard error 31%) to be considered accurate, and the female/moxi combination is too disperse (RSE 56%) to be considered accurate. (The female/moxi error bar actually doesn't even fit on the page!)

## Claims per visit

To examine claims per visit (the upper line), I run $\chi^2$ tests of independence on the 2x2 contingency tables: female vs. male, this drug prescribed vs. not. Nitrofurantoin is definitely distributed differently (minding the caveat mentioned above), and amoxicillin and amox/clav only marginally so (they are the two crosses that could maybe be falling *below* the line of equal consumption per visit; i.e., men getting more of those drugs per visit).

```{r chisq_table}
scientific_round = function(x, digits) {
  if_else(round(x, digits) == 0, sprintf(paste0('%.', as.character(digits), 'g'), x), as.character(round(x, digits)))
}

abx_chisq %>%
  arrange(p.value) %>%
  mutate(p.value.adj=p.adjust(p.value, method='BH')) %>%
  mutate_at(vars(-abx), function(x) scientific_round(x, 3)) %>%
  kable
```

## Claims overall

To evaluate the difference in overall numbers of claims by sex, I use *t*-tests. Again, nitro is the only clear deviation.

```{r ttest_table}
abx_ttest %>%
  arrange(p.value) %>%
  mutate(p.value.adj=p.adjust(p.value, method='BH')) %>%
  mutate_at(vars(-abx), function(x) scientific_round(x, 3)) %>%
  kable
```

@@

Women definitely have more prescriptions overall than men, and definitely more nitrofurantoin prescriptions. However, the difference in nitrofurantoin prescriptions (slightly less than 2 million) is a small fraction of the total difference in prescriptions (about 11 million).

Broadly speaking, it seems like the increased consumption of antibiotics among women is mostly due to increased healthcaure utilization. Unfortunately, NAMCS doesn't have enough data (at least from a single year) to be able to show that any particular drug beyond nitrofurantoin is definitely consumed more among women than men.

Interestingly, there are some hints that amox and amox/clav, though consumed at nearly-equal rates overall, are prescribed at a lower rate *per visit* for women.

The lack of power in this single year motivates either adding more years and/or moving toward Medicare.

>>>>

```{r load_data}
# Caveat: I removed a stratum from the 2011 data that had only one PSU in it.
# Also: I'm not sure that combining years like this is going to be kosher.

years = 2011:2014
n_years = length(years)
mpy = function(x) x/(1e6 * n_years) # millions per year

read_tsv_years = function(fn_template, ...) {
  f = function(y) sprintf(fn_template, y) %>% read_tsv(..., guess_max=1e6) %>% mutate(year=y, visit_id=paste0(y, '_', visit_id))
  lapply(years, f) %>% bind_rows
}

sex_codes = read_tsv('namcs_sex_codes.tsv') %>%
  rename(dx=code)

dx = read_tsv_years('../../db/namcs/data/namcs_dx_%i.tsv') %>%
  left_join(sex_codes, by='dx')

dx_visit = dx %>%
  group_by(visit_id) %>%
  summarize(uti='uti' %in% diagnosis_type,
            acne='acne' %in% diagnosis_type)

top_abx = read_tsv('../../db/namcs/analysis/top_abx.tsv')
abx_cat = read_tsv('../../db/namcs/data/parse/drug_codes/abx_codes.tsv')

drugs = read_tsv_years('../../db/namcs/data/namcs_drugs_%i.tsv') %>%
  mutate(is_abx=cat1 %in% abx_cat$category)

n_abx = drugs %>%
  filter(is_abx) %>%
  group_by(visit_id) %>%
  summarize(n_abx=n())

abx_visit = left_join(top_abx, drugs, by='desc') %>%
  count(visit_id, abx) %>% ungroup() %>%
  spread(abx, n)

visits = read_tsv_years('../../db/namcs/data/namcs_visits_%i.tsv') %>%
  mutate(age_group=case_when(.$age <= 19 ~ 'child',
                             .$age <= 64 ~ 'adult',
                             .$age <= 92 ~ 'elderly')) %>%
  left_join(dx_visit, by='visit_id') %>%
  replace_na(list(uti=FALSE, acne=FALSE)) %>%
  left_join(n_abx, by='visit_id') %>% replace_na(list(n_abx=0)) %>%
  left_join(abx_visit, by='visit_id') %>%
  mutate_at(top_abx$abx, function(x) if_else(is.na(x), 0L, x))

filter_strata = function(df) {
  bad_strata = select(df, survey_id, stratum) %>% distinct() %>% count(stratum) %>% filter(n==1) %$% stratum
  filter(df, !(stratum %in% bad_strata))
}

design_f = function(df) {
  filter_strata(df) %>%
    as.data.frame() %>%
    svydesign(data=., ids=~survey_id, strata=~stratum, weights=~weight, nest=TRUE)
}

design = design_f(visits)
adult_design = subset(design, age_group != 'child')
```

@@

All numbers of prescriptions will be presented in millions per year.

## In children, overall consumption is similar between the sexes

I show total consumption (in millions of prescriptions per year) for the three age groups (children 0-19, adults 20-64, elderly 64-92+). Consumption is similar between the sexes for children. (This probably isn't a surprise, since we don't expect that children have many UTIs or display large differences in healthcare seeking.) Therefore, I'll consider only adults for the rest of this analysis.

```{r children}
svyby(~n_abx, ~sex+age_group, design=design, svytotal) %>%
  mutate_at(vars(n_abx, se), mpy) %>%
  group_by(age_group) %>%
  mutate(odds=n_abx/min(n_abx),
         ci=1.96*se) %>%
  select(age_group, sex, n_abx, `95% CI (one side)`=ci, odds) %>%
  kable
```

## Most-consumed drugs

The common explanation about differences in consumption is that women experience more UTIs than men. We therefore expect that the difference between the sexes will be mostly due to consumption of a few drugs (i.e., ciprofloxacin, TMP/SMX, nitrofurantoin). To simplify analysis, I'll focus on a few of the most-consumed drugs.

The NAMCS data includes information about the therapeutic category of each drug. I looked through those categories and identified the ones that could refer to antibiotics.

```{r abx_cat_table}
abx_cat %>% kable
```

@@

```{r find_top_abx}
# get a list of all the antibiotic names
abx_names = drugs %>%
  semi_join(abx_cat, by=c('cat1'='category')) %$%
  unique(desc)

# for each name, count the visits in which it appears
f = function(name) {
  drugs %>%
    filter(desc==name) %>%
    count(visit_id) %>%
    right_join(select(visits, visit_id, age_group, survey_id, stratum, weight), by='visit_id') %>%
    replace_na(list(n=0)) %>%
    design_f() %>%
    subset(age_group != 'child') %>%
    svytotal(~n, design=.) %>%
    as_data_frame() %>%
    mutate(abx=name) %>%
    select(abx, total, se=n)
}

abx_counts = lapply(abx_names, f) %>% bind_rows()
```

```{r top_abx_table}
abx_counts %>%
  arrange(desc(total)) %>%
  mutate(f=100*round(total/sum(total), 2),
         cum_f_rx=100*round(cumsum(total/sum(total)), 2),
         n=1:nrow(.)) %>%
  select(n, drug=abx, `fraction (%)`=f, `cumulative fraction (%)`=cum_f_rx) %>%
  head(15) %>%
  kable
```

I kept the drugs from azithromycin through clarithromycin, accounting for 80% of all prescriptions.

## Overall healthcare usage by sex

```{r visit_total_odds}
x = svymean(~sex, design=adult_design) %>%
  as_data_frame() %>%
  mutate(sex=c('female', 'male'),
         odds=mean/min(mean))

female_visit_odds = x %>% filter(sex=='female') %$% odds

x %>%
  select(sex, `fraction of visits`=mean, SE, odds) %>%
  kable
```

Overall, more of the visits were made by women. Women make `r round(female_visit_odds - 1, 2)*100`% more visits than men. I'll therefore evaluate consumption by two standards: first, do women consume more antibiotics overall; second, do women consume more antibiotics *per visit*.

## Consumption by drug

```{r models}
abx_diff_f = function(abx) {
  visits %>%
    mutate(sex_abx=case_when(.$sex == 'female' ~ .[[abx]],
                             .$sex == 'male' ~ -.[[abx]])) %>%
    design_f() %>%
    subset(age_group != 'child') %>%
    svyttest(sex_abx~0, design=.) %$%
    data_frame(abx=abx, estimate=estimate, p.value=p.value)
}

abx_total_f = function(abx) {
  svyby(as.formula(paste0('~', abx)), ~sex, adult_design, svytotal) %>%
    rename_(.dots=setNames(list(abx), c('total'))) %>%
    select(sex, total, se) %>%
    mutate(abx=abx)
}

poisson_f = function(abx) {
  svyglm(as.formula(paste0(abx, '~sex')), adult_design) %>%
    tidy %>%
    filter(term != '(Intercept)') %>%
    mutate(abx=abx) %>%
    select(abx, term, estimate, p.value)
}

abx_diff = lapply(top_abx$abx, abx_diff_f) %>% bind_rows()
abx_totals = lapply(top_abx$abx, abx_total_f) %>% bind_rows()
abx_pois = lapply(top_abx$abx, poisson_f) %>% bind_rows()
```

For each drug, I asked how many visits were made, stratified by sex. I show two lines on the graph. The lower line is exactly $y=x$ and shows the line of equal consumption. The upper line has a slope equal to the odds of a visit belong to a female; this is the line of equal consumption per visit.

```{r total_plot}
abx_totals %>%
  mutate_at(vars(total, se), mpy) %>%
  mutate(ci=se*1.96,
         ci_low=total-ci,
         ci_high=total+ci) %>%
  select(abx, sex, total, ci_low, ci_high) %>%
  gather('key', 'value', total:ci_high) %>%
  mutate(key=case_when(.$sex=='male' & .$key=='total' ~ 'x',
                       .$sex=='male' & .$key=='ci_low' ~ 'xmin',
                       .$sex=='male' & .$key=='ci_high' ~ 'xmax',
                       .$sex=='female' & .$key=='total' ~ 'y',
                       .$sex=='female' & .$key=='ci_low' ~ 'ymin',
                       .$sex=='female' & .$key=='ci_high' ~ 'ymax')) %>%
  select(-sex) %>%
  spread(key, value) %>%
  mutate(highlight=abx %in% c('nitro', 'clinda')) %>%
  ggplot(aes(x=x, xmin=xmin, xmax=xmax, y=y, ymin=ymin, ymax=ymax, label=abx, color=highlight)) +
  geom_abline(col='gray') + geom_abline(slope=female_visit_odds, col='gray') +
  geom_point() + geom_errorbar() + geom_errorbarh() +
  scale_color_manual(values=c('black', 'red')) +
  geom_text() +
  #geom_text(nudge_x=1, nudge_y=0.5) +
  #scale_x_continuous(breaks=c(0, 5, 10), limits=c(0, 12)) + xlab('No. rx. for men (millions)') +
  #scale_y_continuous(breaks=c(0, 5, 10), limits=c(0, 14)) + ylab('No. rx. for women (millions)') +
  coord_fixed() +
  theme_minimal()
```

## Prescriptions overall

To evaluate the difference in overall numbers of prescriptions by sex (i.e., deviations from the $y=x$ line in the plot), I performed a one-sample *t*-test on a kludge variable. This variable is the number of antibiotic prescriptions if the visit was made by a woman and *negative* the number of prescriptions if made by a man. The test asks if this value has mean zero.

As we would expect from examining the chart, almost all the drugs have a mean greater than zero, i.e., consumption is greater among women.

```{r abx_diff_table}
scientific_round = function(x, digits=3) {
  if_else(round(x, digits) == 0, sprintf(paste0('%.', as.character(digits), 'g'), x), as.character(round(x, digits)))
}

abx_diff %>%
  mutate(adj.p.value=p.adjust(p.value, method='BH')) %>%
  mutate(sig=case_when(.$adj.p.value < 0.05 ~ '**',
                       .$p.value < 0.05 ~ '*',
                       TRUE ~ '')) %>%
  mutate_if(is.numeric, scientific_round) %>%
  kable
```

## Prescriptions per visit

To examine claims per visit (the upper line), I run Poisson regressions of the number of prescriptions for each drug per visit against sex. The *p*-values here are for the significance of the sex term. Nitro and clinda are enriched among women *per visit*.

```{r pois_table}
abx_pois %>%
  select(abx, estimate, p.value) %>%
  arrange(p.value) %>%
  mutate(p.value.adj=p.adjust(p.value, method='BH')) %>%
  mutate_at(vars(-abx), scientific_round) %>%
  kable
```

## Nitrofurantoin

Nitrofurantoin prescriptions among women are associated with a number of diagnoses, but only two of them---UTI (`5990-`) and dysuria (`7881-`)---meet the basic requirement for data validity (at least 30 counts). Among men, UTI is the leading diagnosis, but it doesn't meet the data validity requirement (there are only 10 records for nitro prescriptions to a male with a UTI diagnosis in the 4 years of data).

## Clindamycin

```{r clinda}
m1 = svyglm(clinda~sex, design=adult_design, family='poisson')
m2 = svyglm(clinda~sex+acne, design=adult_design, family='poisson')
clinda_f_expl = 1 - exp(-m2$coefficients['sexmale']+m1$coefficients['sexmale'])
```

The only clindamycin-associated diagnosis that meets (maybe) the data validity criterion is acne (`7061-`). However, acne explains only a small fraction (`r 100*round(clinda_f_expl, digits=3)`%) of the difference between the sexes.

@@

Women definitely have more prescriptions overall than men, but that consumption seems to be distributed broadly across drugs. For only a few drugs---nitro and clinda---do these data suggest that consumption per visit is greater among women, and those drugs contribute to a minority of the difference in total consumption.
